---
title: "Analyzing a Sample Dataset from the World Bank"
author: "Brian Danielak"
date: "April 12, 2015"
output:
  html_document:
    toc: true
    theme: default
  rmarkdown::tufte_handout:
    toc: true
---

# Introduction

Before we get into analyzing this data, we'll need to do some housekeeping first. In what follows, we'll:

1. Load some non-base R packages we'll be using
2. Load the CSV data into R
3. Check whether the data looks reasonable
4. Perform some data cleaning

## Loading Extra Packages for this Analysis

```{r, message=FALSE}
library(httr)
library(jsonlite)
library(magrittr)
library(readr)
library(dplyr)
library(xtable)
library(knitr)
```



## Getting the Data from the WorldBank

I opted to access the data by [exporting a CSV file][1] from the World Bank's site. 

```{r}
load_data <- function () {
  return (
    read.csv(
      file = "WorldBankData.csv",
      header = TRUE,
      stringsAsFactors = FALSE
    ) %>% 
      tbl_df()
  )
}

world_bank_data <- load_data()
```

## Checking the Reasonability of the Data

### What the data provider says the data should look like

From the [World Bank's website][2]:

> This dataset contains raw response data to a nano-survey that was conducted in Morocco on the right to access to information and the right to petition and motion the government. A nano-survey is an innovative technology that extends a brief survey to a random sampling of internet users. Note: "NA" or "N/A" indicates "No Answer." Sub-national location data is available for 2/3 of survey recipients. 54,441 random internet users in Morocco were exposed to a portion of this survey, with 15,020 respondents providing at least partial responses.

### Checking that the Data are Tabular

We should make sure our data looks OK after all that loading. Sometimes anything from a misplaced comma to a poorly-terminated line can really junk up the works, so let's check by trying to inspect the first few rows.

```{r}
head(world_bank_data)[ ,1:4] %>% 
  kable
```

Good. The tabular structure came through alright. What about the overall structure of the data?

```{r}
str(world_bank_data)
```

Also good! We've 

What about cases?

```{r}
world_bank_data[123, 1:5]
complete.cases(world_bank_data)[122:124]
```





[1]: https://finances.worldbank.org/api/views/tg37-mj88/rows.csv?accessType=DOWNLOAD
[2]: https://finances.worldbank.org/dataset/World-Bank-Morocco-Citizen-Engagement-Nano-Survey-/tg37-mj88


